Spurious Sentience:\\ Dovelating Inflated AI World Model Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}

### 2024-01-25 technical_descriptions.qmd
just a random projection of the training data

## Example: Projecting a World Model onto a latent space

- Take a variational autoencoder (VAE) that was pre-trained on MNIST. 
- Take the [world_place.csv]( https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv) that was used in @gurnee2023language, which maps locations/areas to their latitude and longitude. For each place, it also indicates the corresponding country.
- Take the subset that contains countries that are currently part of the top 10 [FIFA world ranking](https://www.fifa.com/fifa-world-ranking/men?dateId=id14142).
- Assign the current rank to each country, i.e. Argentina gets 1, France gets 2, ...
- Encode all features excluding the FIFA world rank indicator but including `longitude` and `latitude` as continuous variables: $X^{(n \times d)}$ where $n$ is the number of samples and $d$ is the number of resulting features.
- Initialize a small neural network, let's call it a *projector*, that maps from $X$ to a single small hidden layer with sigmoid activation and then form there to the latent space $\mathcal{Z} \subset \mathbb{R}$ of the pre-trained VAE.
- Without performing any training on the *projector*, simply compute a forward pass of the continuously encoded data $X$ through the *projector* and store the output: $\mathbf{Z}^{(n \times k)}$ where $k$ is the dimension of the latent space of the VAE.
- Next, we perform the linear probe on $\mathbf{Z}$ through Ridge regression: $\mathbf{W} = (\mathbf{Z}'\mathbf{Z} + \lambda \mathbf{I}) (\mathbf{Z}'\mathbf{Y})^{-1}$ where $\mathbf{Y}$ is the $(n \times 2)$ matrix containing the longitude and latitude for each sample. 
- Finally, compute the predicted coordinate for each samples as $\widehat{\mathbf{Y}}=\mathbf{Z}\mathbf{W}$
- Plot the results on a world map and be amazed.

> I think what's going on here is that we essentially randomly project a small set of features into a higher-dimensional space ($d < k$). The linear probe is then having an easy time separating features in the high-dimensional space, even though the projection is completely random. Even though one might be momentarily amazed by this, there is really nothing surprising at all going on here. We can still draw a very interesting parallel to @gurnee2023language: we know that geographical coordinates are in the training corpus of Llama-2; that information gets passed through a massive latent space (the model); they run probes on a subsapce of the model (still massive) and manage to linearly separate their variable of interest (coordinates). I don't know how accurate that parallel really is but it just seems striking to me that ultimately what they is regressing a 2-dimensional output variable on this massive feature matrix ("ranging from 4,096 to 8,192 features" for roughly 40,000 total samples, can't figure out how big the train/test sets are, respectively.). I'm not good enough at theory but maybe there's really just no surprise at all here?

>>>>

## Example: Random Projections of World Models

- Take the [world_place.csv]( https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv) that was used in @gurnee2023language, which maps locations/areas to their latitude and longitude. For each place, it also indicates the corresponding country.
- Take the subset that contains countries that are currently part of the top 10 [FIFA world ranking](https://www.fifa.com/fifa-world-ranking/men?dateId=id14142).
- Assign the current rank to each country, i.e. Argentina gets 1, France gets 2, ...
- Encode all features excluding the FIFA world rank indicator but including `longitude` and `latitude` as continuous variables: $X^{(n \times d)}$ where $n$ is the number of samples and $d$ is the number of resulting features.
- Initialize a small neural network, let's call it a *projector*, that maps from $X$ to a single small hidden layer with sigmoid activation and then from there to a lower-dimensional output space $\mathcal{Z} \subset \mathbb{R}$.
- Without performing any training on the *projector*, simply compute a forward pass of the continuously encoded data $X$ through the *projector* and store the output: $\mathbf{Z}^{(n \times k)}$ where $k$ is the dimension of the latent space of the VAE.
- Next, we perform the linear probe on $\mathbf{Z}$ through Ridge regression: $\mathbf{W} = (\mathbf{Z}'\mathbf{Z} + \lambda \mathbf{I}) (\mathbf{Z}'\mathbf{Y})^{-1}$ where $\mathbf{Y}$ is the $(n \times 2)$ matrix containing the longitude and latitude for each sample. 
- Finally, compute the predicted coordinate for each samples as $\widehat{\mathbf{Y}}=\mathbf{Z}\mathbf{W}$
- Plot the results on a world map and be amazed.

> We can draw a very interesting parallel to @gurnee2023language: we know that geographical coordinates are in the training corpus of Llama-2; that information gets passed through a massive latent space (the model); they run probes on a subsapce of the model (still massive) and manage to linearly separate their variable of interest (coordinates). I don't know how accurate that parallel really is but it just seems striking to me that ultimately what they is regressing a 2-dimensional output variable on this massive feature matrix ("ranging from 4,096 to 8,192 features" for roughly 40,000 total samples, can't figure out how big the train/test sets are, respectively.).

### 2024-01-25 technical_descriptions.tex
just a random projection of the training data

\begin{document}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, frame hidden]}{\end{tcolorbox}}\fi

\hypertarget{example-projecting-a-world-model-onto-a-latent-space}{%
\subsection{Example: Projecting a World Model onto a latent
space}\label{example-projecting-a-world-model-onto-a-latent-space}}

\begin{itemize}
\tightlist
\item
  Take a variational autoencoder (VAE) that was pre-trained on MNIST.
\item
  Take the
  \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv}
  that was used in Gurnee and Tegmark (2023), which maps locations/areas
  to their latitude and longitude. For each place, it also indicates the
  corresponding country.
\item
  Take the subset that contains countries that are currently part of the
  top 10
  \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA
  world ranking}.
\item
  Assign the current rank to each country, i.e.~Argentina gets 1, France
  gets 2, \ldots{}
\item
  Encode all features excluding the FIFA world rank indicator but
  including \texttt{longitude} and \texttt{latitude} as continuous
  variables: \(X^{(n \times d)}\) where \(n\) is the number of samples
  and \(d\) is the number of resulting features.
\item
  Initialize a small neural network, let's call it a \emph{projector},
  that maps from \(X\) to a single small hidden layer with sigmoid
  activation and then form there to the latent space
  \(\mathcal{Z} \subset \mathbb{R}\) of the pre-trained VAE.
\item
  Without performing any training on the \emph{projector}, simply
  compute a forward pass of the continuously encoded data \(X\) through
  the \emph{projector} and store the output:
  \(\mathbf{Z}^{(n \times k)}\) where \(k\) is the dimension of the
  latent space of the VAE.
\item
  Next, we perform the linear probe on \(\mathbf{Z}\) through Ridge
  regression:
  \(\mathbf{W} = (\mathbf{Z}'\mathbf{Z} + \lambda \mathbf{I}) (\mathbf{Z}'\mathbf{Y})^{-1}\)
  where \(\mathbf{Y}\) is the \((n \times 2)\) matrix containing the
  longitude and latitude for each sample.
\item
  Finally, compute the predicted coordinate for each samples as
  \(\widehat{\mathbf{Y}}=\mathbf{Z}\mathbf{W}\)
\item
  Plot the results on a world map and be amazed.
\end{itemize}

\begin{quote}
I think what's going on here is that we essentially randomly project a
small set of features into a higher-dimensional space (\(d < k\)). The
linear probe is then having an easy time separating features in the
high-dimensional space, even though the projection is completely random.
Even though one might be momentarily amazed by this, there is really
nothing surprising at all going on here. We can still draw a very
interesting parallel to Gurnee and Tegmark (2023): we know that
geographical coordinates are in the training corpus of Llama-2; that
information gets passed through a massive latent space (the model); they
run probes on a subsapce of the model (still massive) and manage to
linearly separate their variable of interest (coordinates). I don't know
how accurate that parallel really is but it just seems striking to me
that ultimately what they is regressing a 2-dimensional output variable
on this massive feature matrix (``ranging from 4,096 to 8,192 features''
for roughly 40,000 total samples, can't figure out how big the
train/test sets are, respectively.). I'm not good enough at theory but
maybe there's really just no surprise at all here?
\end{quote}

>>>>

\begin{document}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, sharp corners, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi

\hypertarget{example-random-projections-of-world-models}{%
\subsection{Example: Random Projections of World
Models}\label{example-random-projections-of-world-models}}

\begin{itemize}
\tightlist
\item
  Take the
  \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv}
  that was used in Gurnee and Tegmark (2023), which maps locations/areas
  to their latitude and longitude. For each place, it also indicates the
  corresponding country.
\item
  Take the subset that contains countries that are currently part of the
  top 10
  \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA
  world ranking}.
\item
  Assign the current rank to each country, i.e.~Argentina gets 1, France
  gets 2, \ldots{}
\item
  Encode all features excluding the FIFA world rank indicator but
  including \texttt{longitude} and \texttt{latitude} as continuous
  variables: \(X^{(n \times d)}\) where \(n\) is the number of samples
  and \(d\) is the number of resulting features.
\item
  Initialize a small neural network, let's call it a \emph{projector},
  that maps from \(X\) to a single small hidden layer with sigmoid
  activation and then from there to a lower-dimensional output space
  \(\mathcal{Z} \subset \mathbb{R}\).
\item
  Without performing any training on the \emph{projector}, simply
  compute a forward pass of the continuously encoded data \(X\) through
  the \emph{projector} and store the output:
  \(\mathbf{Z}^{(n \times k)}\) where \(k\) is the dimension of the
  latent space of the VAE.
\item
  Next, we perform the linear probe on \(\mathbf{Z}\) through Ridge
  regression:
  \(\mathbf{W} = (\mathbf{Z}'\mathbf{Z} + \lambda \mathbf{I}) (\mathbf{Z}'\mathbf{Y})^{-1}\)
  where \(\mathbf{Y}\) is the \((n \times 2)\) matrix containing the
  longitude and latitude for each sample.
\item
  Finally, compute the predicted coordinate for each samples as
  \(\widehat{\mathbf{Y}}=\mathbf{Z}\mathbf{W}\)
\item
  Plot the results on a world map and be amazed.
\end{itemize}

\begin{quote}
We can draw a very interesting parallel to Gurnee and Tegmark (2023): we
know that geographical coordinates are in the training corpus of
Llama-2; that information gets passed through a massive latent space
(the model); they run probes on a subsapce of the model (still massive)
and manage to linearly separate their variable of interest
(coordinates). I don't know how accurate that parallel really is but it
just seems striking to me that ultimately what they is regressing a
2-dimensional output variable on this massive feature matrix (``ranging
from 4,096 to 8,192 features'' for roughly 40,000 total samples, can't
figure out how big the train/test sets are, respectively.).
\end{quote}
